---
title: "US birth oblig"
author: "Morten S Østrem"
date: "11/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = FALSE)
library(tidyverse)
library(lubridate)
```

## US Births by Sex, visualized

This document intends to depict a visual representation of statistics on newborns in the US. Primarily, the ratios between male and female newborns will be examined.

# Proportion of Boys to Girls

The following graph compares the total number of cumulative births for each sex over the time period of a year in 2017, 2018 and 2019, respectively. As we can see, the number of male newborns is slightly higher than that of females.

```{r proportion, include=FALSE, echo=FALSE, results='hide', warning=FALSE}
df_2017_raw <- read_fwf(
  file='Nat2017PublicUS.c20180516.r20180808.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2017_sex <- df_2017_raw %>% 
  select(X2, X3, X7) %>% 
  rename(., Year = X2, Month = X3,Sex = X7) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Date = floor_date(Date, "month")) %>% 
  summarise(n = cumsum(n())) %>% 
  mutate("Cumulative" = cumsum(n))

rm(df_2017_raw)


df_2018_raw <- read_fwf(
  file='Nat2018PublicUS.c20190509.r20190717.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2018_sex <- df_2018_raw %>% 
  select(X2, X3, X7) %>% 
  rename(., Year = X2, Month = X3,Sex = X7) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Date = floor_date(Date, "month")) %>% 
  summarise(n = cumsum(n())) %>% 
  mutate("Cumulative" = cumsum(n))

rm(df_2018_raw)


df_2019_raw <- read_fwf(
  file='Nat2019PublicUS.c20200506.r20200915.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2019_sex <- df_2019_raw %>% 
  select(X2, X3, X7) %>% 
  rename(., Year = X2, Month = X3,Sex = X7) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Date = floor_date(Date, "month")) %>% 
  summarise(n = cumsum(n())) %>% 
  mutate("Cumulative" = cumsum(n))

rm(df_2019_raw)

df_total_sex <- rbind(df_2017_sex, df_2018_sex, df_2019_sex) %>% 
  mutate(Year = factor(year(Date)), Month = month(Date)) %>% 
  select(Sex, Cumulative, Month, Year)


```

```{r propplot, include=TRUE, echo=FALSE, results='hide', warning=FALSE}

ggplot(df_total_sex, aes(Month, Cumulative, group = Sex, color = Sex)) +
  geom_line() +
  facet_wrap(~Year) +
  xlab("Month (Jan - Dec)") +
  ylab("Cumulative Births by Sex (M/F)") +
  labs(color = "Sex") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text=element_text(size=11, face="bold"),
        axis.title=element_text(size=14, face="bold"))


# Cleanup
rm(df_2017_sex, df_2018_sex, df_2019_sex, df_total_sex)
```

Continuing, we attempt to visualize the relationship between male and female newborns with respect to weight in grams. The average weight of newborn males seems to lie somewhere around 100g higher than that of females.

```{r weight, include=FALSE, echo=FALSE, results='hide', warning=FALSE}
df_2017_raw <- read_fwf(
  file='Nat2017PublicUS.c20180516.r20180808.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2017_weight <- df_2017_raw %>% 
  select(X2, X3, X7, X9) %>% 
  rename(., Year = X2, Month = X3, Sex = X7, Weight = X9) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex), Weight = as.numeric(Weight)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Date = floor_date(Date, "month")) %>% 
  summarise("avg_weight" = mean(Weight))

rm(df_2017_raw)


df_2018_raw <- read_fwf(
  file='Nat2018PublicUS.c20190509.r20190717.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2018_weight <- df_2018_raw %>% 
  select(X2, X3, X7, X9) %>% 
  rename(., Year = X2, Month = X3, Sex = X7, Weight = X9) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex), Weight = as.numeric(Weight)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Date = floor_date(Date, "month")) %>% 
  summarise("avg_weight" = mean(Weight))

rm(df_2018_raw)


df_2019_raw <- read_fwf(
  file='Nat2019PublicUS.c20200506.r20200915.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2019_weight <- df_2019_raw %>% 
  select(X2, X3, X7, X9) %>% 
  rename(., Year = X2, Month = X3, Sex = X7, Weight = X9) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex), Weight = as.numeric(Weight)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Date = floor_date(Date, "month")) %>% 
  summarise("avg_weight" = mean(Weight))

rm(df_2019_raw)

df_total_weight <- rbind(df_2017_weight, df_2018_weight, df_2019_weight) %>% 
  mutate(Year = factor(year(Date)), Month = month(Date)) %>% 
  select(Sex, avg_weight, Month, Year)

```

```{r weightplot, include=TRUE, echo=FALSE, results='hide', warning=FALSE}

ggplot(df_total_weight, aes(Month, avg_weight, group = Sex, color = Sex)) +
  geom_line() +
  facet_wrap(~Year) +
  xlab("Month (Jan - Dec)") +
  ylab("Average Birth Weight (g)") +
  ylim(3100, 3400) +
  labs(color = "Sex") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text=element_text(size=11, face="bold"),
        axis.title=element_text(size=14, face="bold"))

# Cleanup
rm(df_2017_weight, df_2018_weight, df_2019_weight, df_total_weight)
```

Finally, we examine the ratio of boys to girls by day of birth in the week.

```{r weekday, include=FALSE, echo=FALSE, results='hide', warning=FALSE}
df_2017_raw <- read_fwf(
  file='Nat2017PublicUS.c20180516.r20180808.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2017_weekday <- df_2017_raw %>% 
  select(X2, X3, X5, X7) %>% 
  rename(., Year = X2, Month = X3, Weekday = X5, Sex = X7) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Weekday, Date = floor_date(Date, "month")) %>% 
  summarise(n = cumsum(n())) %>% 
  mutate("Cumulative" = cumsum(n))

rm(df_2017_raw)


df_2018_raw <- read_fwf(
  file='Nat2018PublicUS.c20190509.r20190717.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2018_weekday <- df_2018_raw %>% 
  select(X2, X3, X5, X7) %>% 
  rename(., Year = X2, Month = X3, Weekday = X5, Sex = X7) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Weekday, Date = floor_date(Date, "month")) %>% 
  summarise(n = cumsum(n())) %>% 
  mutate("Cumulative" = cumsum(n))

rm(df_2018_raw)


df_2019_raw <- read_fwf(
  file='Nat2019PublicUS.c20200506.r20200915.txt',
  # Fill, Year, Mo, Time, Day, Fill, Sex, Fill, Weight
  fwf_widths(c(8, 4, 2, 8, 1, 451, 1, 28, 4))
)

df_2019_weekday <- df_2019_raw %>% 
  select(X2, X3, X5, X7) %>% 
  rename(., Year = X2, Month = X3, Weekday = X5, Sex = X7) %>% 
  mutate(Month = as.numeric(Month), Sex = as.factor(Sex)) %>% 
  mutate(Date = dmy(paste(1, .$Month, .$Year, sep="-"))) %>% 
  group_by(Sex, Weekday, Date = floor_date(Date, "month")) %>% 
  summarise(n = cumsum(n())) %>% 
  mutate("Cumulative" = cumsum(n))


rm(df_2019_raw)


df_total_weekday <- rbind(df_2017_weekday, df_2018_weekday, df_2019_weekday) %>% 
  mutate(Year = factor(year(Date)), Month = month(Date)) %>% 
  select(Sex, Month, Year, Weekday, Cumulative)

```

```{r dayplot, include=TRUE, echo=FALSE, results='hide', warning=FALSE, fig.height=100}

ggplot(df_total_weekday, aes(Month, Cumulative, group = Sex, color = Sex)) +
  geom_line() +
  facet_grid(Weekday ~ Year) +
  xlab("Month (Jan - Dec)") +
  ylab("Cumulative Births by Sex (M/F), by Weekdays (Mon-Sun)") +
  labs(color = "Sex") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.text=element_text(size=11, face="bold"),
        axis.title=element_text(size=14, face="bold"))

rm(df_2017_weekday, df_2018_weekday, df_2019_weekday, df_total_weekday)
```


